# Expected Utility Theory

Under expected utility theory, people do not seek to maximize expected value but instead, maximize expected utility.

Under expected utility theory, people choose between risky prospects (prospect being another word for lottery or gamble common in the literature) by comparing expected utility values. An agent would pick the option that maximises their expected utility.

## Calculating expected utility theory

Consider a prospect with final outcomes $x_1, ..., x_n$, with each outcome occurring with probability $p_1, ..., p_n$. Each outcome would deliver utility $U(x_i)$.

Expected utility, $E[U(X)]$, is calculated using the following formula:

\begin{align*}
E[U(X)]&=p_1U(x_1)+p_2U(x_2)+ ... +p_nU(x_n) \\[6pt]
&=\sum_{i=1}^n p_iU(x_i)
\end{align*}

You can think of this formula as comprising the following steps:

1. Define utility $U(x_i)$ over final outcomes $x_1, ..., x_n$

2. Weight the utility of each outcome $U(x_i)$ by the probability $p_i$ of outcome $x_i$

3. Add the weighted utilities.

There is an important note to make here regarding outcomes $x_1, ..., x_n$. Typically, these outcomes are not just the payoffs from the gamble, but rather the agent's final position. If the agent has wealth of \$100 and is offered a coin flip to win or lose \$10, the outcomes are typically taken to be \$90 and \$110. Their decision depends on their current wealth. As a result, expected utility is often represented as in this equation:

\begin{align*}
E[U(W+X)]&=p_1U(W+x_1)+p_2U(W+x_2)  \\[6pt]
&\qquad +...+p_nU(W+x_n) \\[6pt]
&=\sum_{i=1}^n p_iU(W+x_i)
\end{align*}

## Attitudes toward risk

Expected utility theory allows us to examine an agent's attitude toward risk.

There are three possible attitudes toward risk: risk aversion, risk neutrality, and risk seeking.

If a person prefers a sure amount to a gamble with the same expected value, they are risk averse. That is, the utility of the expected value of $X$ is greater than the expected utility of $X$.

$$
U(E[X])>E[U(X)]
$$

If a person prefers a gamble to a sure amount of the same expected value, they are risk-seeking. That is, the expected utility of $X$ is greater than the utility of the expected value of $X$.

$$
U(E[X])<E[U(X)]
$$

If a person is indifferent between a gamble and receiving the expected value of the gamble with certainty, they are risk neutral. That is. the utility of the expected value of $X$ is equal to the expected utility of $X$.

$$
U(E[X])=E[U(X)]
$$

### Certainty equivalent

One useful concept in the analysis of attitudes to risk is the certainty equivalent. The certainty equivalent (CE) of a gamble X is the amount of money such that you are indifferent between taking the gamble and taking the money. That is, the utility of the certainty equivalent is equal to the expected utility of the gamble.

$$
u(CE)=E[U(X)]
$$

For a risk-averse person, the certainty equivalent of the bet is less than the expected value of the gamble. The certainty equivalent is equal to the expected value in the case of risk neutrality. A risk-seeking person would have a certainty equivalent higher than the expected value of the gamble.

I will now look at each of the attitudes in turn.

### Risk aversion

A risk-averse person prefers a sure amount to a gamble with the same expected value. If I strongly prefer \$10 for certain to a gamble with an expected value of \$10, I am risk averse. The certainty equivalent of the prospect for this person would be less than \$10.

The following chart illustrates. The x-axis is the amount of the good over which the person receives utility. In the case of monetary gambles of the type we are talking about, the x-axis is the amount of money. The y-axis is the utility the person receives from that money.

The utility curve is a plot of the utility function for each amount of money.

The gamble shown in this chart has two possible outcomes, $x_1$ and $x_2$. An outcome of $x_1$ would result in utility $U(x_1)$. An outcome of $x_2$ would result in utility $U(x_2)$.

I have drawn a straight dash-dot line between the points on the utility curve for each of those outcomes. The expected utility from any gamble involving those two outcomes would fall on that line. Where on that line would depend on the probability of each outcome and it would occur at a point in line with the expected value of the gamble. You can see the expected value of the gamble marked on the x-axis. If we extend up from that point we can read the expected utility of the gamble, E[U(X)], from the y-axis.

One way to think about why projecting the expected value onto the expected utility line identifies the expected utility of the bet is that each of the expected value and expected utility are weighted by the same probabilities. They both lie the same horizontal distance between the two potential outcomes.

In this particular example, the line between the utility of the two outcomes is always below the utility curve. That is, for any probabilities involving those two outcomes (except one of those outcomes with certainty), the expected utility of the prospect is less than the utility of the expected value of the prospect. In mathematical terms:

$$
U(E[x])>E[U(x)]
$$

```{r}
#| label: fig-risk-aversion
#| fig-cap: "Risk aversion"

library(ggplot2)

u <- function(x){
  log(x)
}

df <- data.frame(
  x=seq(1,220,0.1),
  y=NA
)

df$y <- u(df$x)

#Variables for plot (may not match labels as not done to scale)
#Payoffs from gamble
p1 <- 0.5
x1 <- 30 #loss
p2 <- 0.5
x2 <- 200 #win
ev <- p1*x1+p2*x2 #expected value of gamble
xc <-115 #certain outcome
ce <-(p1*u(x1)+p2*u(x2))^(1/1.5) #certainty equivalent
px2 <-(ev-x1)/(x2-x1)

ggplot(mapping = aes(x, y)) +

  #Plot the utility curve
  geom_line(data = df) +
  geom_vline(xintercept = 0, linewidth=0.25)+ 
  geom_hline(yintercept = 0, linewidth=0.25)+
  labs(x = "x", y = "U(x)")+

  # Set the theme
  theme_minimal()+

  #remove numbers on each axis
  theme(axis.text.x = element_blank(),
            axis.text.y = element_blank(),
            axis.title=element_text(size=14,face="bold"),
            axis.title.y = element_text(angle=0, vjust=0.5))+

  #set limits - need to include room for labels
  coord_cartesian(xlim = c(-25, 220), ylim = c(-0.25, 6))+

  #Add labels x1, U(x1) and line to curve indicating each
  annotate("text", x = x1, y = 0, label = "x1", size = 4, hjust = 0.5, vjust = 1.5)+
  annotate("segment", x = x1, y = 0, xend = x1, yend = u(x1), linewidth = 0.5, colour = "black", linetype="dotted")+
  annotate("segment", x = 0, y = u(x1), xend = x1, yend = u(x1), linewidth = 0.5, colour = "black", linetype="dotted")+
  annotate("text", x = 0, y = u(x1), label = "U(x1)", size = 4, hjust = 1.05, vjust = 0.6)+

  #Add line to curve indicating utility of expected value
  annotate("segment", x = xc, y = 0, xend = xc, yend = u(xc), linewidth = 0.5, colour = "black", linetype="dotted")+
  annotate("segment", x = 0, y = u(xc), xend = xc, yend = u(xc), linewidth = 0.5, colour = "black", linetype="dotted")+
  annotate("text", x = 0, y = u(xc), label = "U[E(X)]", size = 4, hjust = 1.05, vjust = 0.3)+

  #Add expected utility line
  annotate("segment", x = x1, xend = x2, y = u(x1), yend = u(x2), linewidth = 0.5, colour = "black", linetype="dotdash")+

  #Add labels x2, U(x2) and line to curve indicating each
  annotate("text", x = x2, y = 0, label = "x2", size = 4, hjust = 0.4, vjust = 1.5)+
  annotate("segment", x = x2, y = 0, xend = x2, yend = u(x2), linewidth = 0.5, colour = "black", linetype="dotted")+
  annotate("segment", x = 0, y = u(x2), xend = x2, yend = u(x2), linewidth = 0.5, colour = "black", linetype="dotted")+
  annotate("text", x = 0, y = u(x2), label = "U(x2)", size = 4, hjust = 1.05, vjust = 0.45)+

  #Add labels E[X], E[U(X)] and curve indicating each
  annotate("text", x = ev, y = 0, label = "E[X]", size = 4, hjust = 0.4, vjust = 1.5)+
  annotate("segment", x = ev, y = 0, xend = ev, yend = u(x1)+(u(x2)-u(x1))*px2, linewidth = 0.5, colour = "black", linetype="dashed")+
  annotate("segment", x = 0, y = u(x1)+(u(x2)-u(x1))*px2, xend = ev, yend = u(x1)+(u(x2)-u(x1))*px2, linewidth = 0.5, colour = "black", linetype="dashed")+
  annotate("text", x = 0, y = u(x1)+(u(x2)-u(x1))*px2, label = "E[U(X)]", size = 4, hjust = 1.05, vjust = 0.45)+

  #Add vertical line indicating certainty equivalent and labelled "CE"
  annotate("segment", x = ce, xend = ce, y = 0, yend = u(ce), linewidth = 0.5, colour = "black", linetype="dotted")+
  annotate("text", x = ce, y = 0, label = "CE", size = 4, hjust = 0.4, vjust = 1.5)

```

The expected utility line between the two outcomes is always below the the utility curve as the curve is concave. A concave curve leads to risk aversion.

Finally, the horizontal dashed line identifying E[U(X)] allows us to identify the certainty equivalent of the gamble. At the point on the x-axis marked CE, the utility from CE is equal to the expected utility of the prospect with expected value E[X]. As the certainty equivalent is less than the expected value, this provides another way of saying that the person is risk averse.

### Risk neutrality

A risk-neutral person is an expected value maximiser. They are indifferent between \$10 for certain and a gamble with an expected value of \$10. The certainty equivalent of the prospect for this person would also be \$10.

The following chart illustrates. Again we have two possible outcomes, $x_1$ and $x_2$ with resulting utility $U(x_1)$ and $U(x_2)$.

A line between the points on the utility curve for each of those outcomes lies on the utility curve itself. For any probability, the utility of the expected value and the expected utility are the same.

```{r}
#| label: fig-risk-neutral
#| fig-cap: "Risk neutral"

library(ggplot2)

u <- function(x){
  x
}

df <- data.frame(
  x=seq(1,220,0.1),
  y=NA
)

df$y <- u(df$x)

#Variables for plot (may not match labels as not done to scale)
#Payoffs from gamble
p1 <- 0.5
x1 <- 30 #loss
p2 <- 0.5
x2 <- 200 #win
ev <- p1*x1+p2*x2 #expected value of gamble
xc <-115 #certain outcome
ce <-p1*u(x1)+p2*u(x2) #certainty equivalent
px2 <-(ev-x1)/(x2-x1)

ggplot(mapping = aes(x, y)) +

  #Plot the utility curve
  geom_line(data = df) +
  geom_vline(xintercept = 0, linewidth=0.25)+ 
  geom_hline(yintercept = 0, linewidth=0.25)+
  labs(x = "x", y = "U(x)")+

  # Set the theme
  theme_minimal()+

  #remove numbers on each axis
  theme(axis.text.x = element_blank(),
            axis.text.y = element_blank(),
            axis.title=element_text(size=14,face="bold"),
            axis.title.y = element_text(angle=0, vjust=0.5))+

  #set limits - need to include room for labels
  coord_cartesian(xlim = c(-40, 220), ylim = c(-0.25, 220))+

  #Add labels x1, U(x1) and line to curve indicating each
  annotate("text", x = x1, y = 0, label = "x1", size = 4, hjust = 0.5, vjust = 1.5)+
  annotate("segment", x = x1, y = 0, xend = x1, yend = u(x1), linewidth = 0.5, colour = "black", linetype="dotted")+
  annotate("segment", x = 0, y = u(x1), xend = x1, yend = u(x1), linewidth = 0.5, colour = "black", linetype="dotted")+
  annotate("text", x = 0, y = u(x1), label = "U(x1)", size = 4, hjust = 1.05, vjust = 0.6)+

  #Add line to curve indicating utility of expected value
  annotate("segment", x = xc, y = 0, xend = xc, yend = u(xc), linewidth = 0.5, colour = "black", linetype="dotted")+
  annotate("segment", x = 0, y = u(xc), xend = xc, yend = u(xc), linewidth = 0.5, colour = "black", linetype="dotted")+
  annotate("text", x = 0, y = u(xc), label = "U[E(X)]=E[U(X)]", size = 4, hjust = 1.05, vjust = 0.3)+

  #Add labels x2, U(x2) and line to curve indicating each
  annotate("text", x = x2, y = 0, label = "x2", size = 4, hjust = 0.4, vjust = 1.5)+
  annotate("segment", x = x2, y = 0, xend = x2, yend = u(x2), linewidth = 0.5, colour = "black", linetype="dotted")+
  annotate("segment", x = 0, y = u(x2), xend = x2, yend = u(x2), linewidth = 0.5, colour = "black", linetype="dotted")+
  annotate("text", x = 0, y = u(x2), label = "U(x2)", size = 4, hjust = 1.05, vjust = 0.45)+

  #Add labels E[X], E[U(X)] and curve indicating each
  annotate("text", x = ev, y = 0, label = "E[X]=CE", size = 4, hjust = 0.4, vjust = 1.5)

```

### Risk seeking

A risk-seeking person prefers a gamble to a sure sum equal to the expected value of that gamble. The certainty equivalent is more than the expected value of the gamble. The gamble has value in and of itself.

The following chart illustrates. Again I have drawn a dash-dot-dot line between the points on the utility curve for each of those outcomes. That line is always above the utility curve. That is, for any probabilities involving those two outcomes (except one of those outcomes with certainty), the expected utility of the prospect is more than the utility of the expected value of the prospect.

```{r}
#| label: fig-risk-seeking
#| fig-cap: "Risk seeking"

library(ggplot2)

u <- function(x){
  x^1.5
}

df <- data.frame(
  x=seq(1,220,0.1),
  y=NA
)

df$y <- u(df$x)

#Variables for plot (may not match labels as not done to scale)
#Payoffs from gamble
p1 <- 0.5
x1 <- 30 #loss
p2 <- 0.5
x2 <- 200 #win
ev <- p1*x1+p2*x2 #expected value of gamble
xc <-115 #certain outcome
ce <-(p1*u(x1)+p2*u(x2))^(1/1.5) #certainty equivalent
px2 <-(ev-x1)/(x2-x1)

ggplot(mapping = aes(x, y)) +

  #Plot the utility curve
  geom_line(data = df) +
  geom_vline(xintercept = 0, linewidth=0.25)+ 
  geom_hline(yintercept = 0, linewidth=0.25)+
  labs(x = "x", y = "U(x)")+

  # Set the theme
  theme_minimal()+

  #remove numbers on each axis
  theme(axis.text.x = element_blank(),
            axis.text.y = element_blank(),
            axis.title=element_text(size=14,face="bold"),
            axis.title.y = element_text(angle=0, vjust=0.5))+

  #set limits - need to include room for labels
  coord_cartesian(xlim = c(-25, 220), ylim = c(-0.25, 3000))+

  #Add labels x1, U(x1) and line to curve indicating each
  annotate("text", x = x1, y = 0, label = "x1", size = 4, hjust = 0.5, vjust = 1.5)+
  annotate("segment", x = x1, y = 0, xend = x1, yend = u(x1), linewidth = 0.5, colour = "black", linetype="dotted")+
  annotate("segment", x = 0, y = u(x1), xend = x1, yend = u(x1), linewidth = 0.5, colour = "black", linetype="dotted")+
  annotate("text", x = 0, y = u(x1), label = "U(x1)", size = 4, hjust = 1.05, vjust = 0.6)+

  #Add line to curve indicating utility of expected value
  annotate("segment", x = xc, y = 0, xend = xc, yend = u(xc), linewidth = 0.5, colour = "black", linetype="dotted")+
  annotate("segment", x = 0, y = u(xc), xend = xc, yend = u(xc), linewidth = 0.5, colour = "black", linetype="dotted")+
  annotate("text", x = 0, y = u(xc), label = "U[E(X)]", size = 4, hjust = 1.05, vjust = 0.3)+

  #Add expected utility line
  annotate("segment", x = x1, xend = x2, y = u(x1), yend = u(x2), linewidth = 0.5, colour = "black", linetype="dotdash")+

  #Add labels x2, U(x2) and line to curve indicating each
  annotate("text", x = x2, y = 0, label = "x2", size = 4, hjust = 0.4, vjust = 1.5)+
  annotate("segment", x = x2, y = 0, xend = x2, yend = u(x2), linewidth = 0.5, colour = "black", linetype="dotted")+
  annotate("segment", x = 0, y = u(x2), xend = x2, yend = u(x2), linewidth = 0.5, colour = "black", linetype="dotted")+
  annotate("text", x = 0, y = u(x2), label = "U(x2)", size = 4, hjust = 1.05, vjust = 0.45)+

  #Add labels E[X], E[U(X)] and curve indicating each
  annotate("text", x = ev, y = 0, label = "E[X]", size = 4, hjust = 0.4, vjust = 1.5)+
  annotate("segment", x = ev, y = 0, xend = ev, yend = u(x1)+(u(x2)-u(x1))*px2, linewidth = 0.5, colour = "black", linetype="dashed")+
  annotate("segment", x = 0, y = u(x1)+(u(x2)-u(x1))*px2, xend = ev, yend = u(x1)+(u(x2)-u(x1))*px2, linewidth = 0.5, colour = "black", linetype="dashed")+
  annotate("text", x = 0, y = u(x1)+(u(x2)-u(x1))*px2, label = "E[U(X)]", size = 4, hjust = 1.05, vjust = 0.45)+

  #Add vertical line indicating certainty equivalent and labelled "CE" plus line extending to utility curve
  annotate("segment", x = ce, xend = ce, y = 0, yend = u(ce), linewidth = 0.5, colour = "black", linetype="dotted")+
  annotate("text", x = ce, y = 0, label = "CE", size = 4, hjust = 0.4, vjust = 1.5)+
  annotate("segment", x = 0, y = u(ce), xend = ce, yend = u(ce), linewidth = 0.5, colour = "black", linetype="dotted")

```

The expected utility line between the two outcomes is always above the the utility curve as the curve is convex. A convex curve leads to risk seeking behaviour.
